---
title: "ps5-rmarkdown"
author: "Mira Nair"
date: "2023-02-19"
output: html_document
---

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
```{r, echo=FALSE}
library(tidyverse)
library(dplyr)
gapminder <- read_delim("gapminder.csv.bz2")
```

There are `r ncol(gapminder)` columns and `r nrow(gapminder)` rows.
Example of the dataset: 
```{r , echo=FALSE}
head(gapminder, 3)
```
The sample dataset looks good!

1. There are `r n_distinct(gapminder$name)` countries under "name", there are `r n_distinct(gapminder$iso2)` countries under "iso2", and there are `r n_distinct(gapminder$iso3)` under "iso3".

2. 
```{r, echo=FALSE}
gapminder %>%
  group_by(iso2)%>%
  summarise(n = n_distinct(name)) %>%
            filter(min_rank(desc(n))<6)
            
```

```{r, echo=FALSE}
gapminder %>%
  filter(is.na(iso2))%>%
  distinct(name)
```


```{r, echo=FALSE}
gapminder %>%
  group_by(iso3)%>%
  summarise(n = n_distinct(name)) %>%
            filter(min_rank(desc(n))<6)
            
```

```{r, echo=FALSE}
gapminder %>%
  filter(nchar(iso3) > 3)%>%
  distinct(name)
```

3. These are the earliest and latest dates in the dataset.
```{r, echo=FALSE}
new1 <- na.omit(gapminder$time)
min(new1)
max(new1)
```




--->3 CO2 emissions
1. This the table for all of the missing co2_PC emissions throughout the years and the first year, 1960, has the most missing data.
```{r, echo=FALSE}
gapminder%>%
  group_by(time)%>%
  filter(is.na(co2_PC))%>%
  count(is.na(co2_PC))
```

 
2. This is a plot of CO2 emissions in the U.S, China, India, Canada, and Mexico over time.
```{r, echo=FALSE}
new <- filter(gapminder, name == 'China'| name == 'India' | name == 'U.S' | name == 'Mexico' | name == 'Canada' | name == 'United States of America')

ggplot(new) +
  geom_point(mapping = aes(x=time, y=co2, color = name))
```
China has the highest CO2 emissions out of all of the countries. America also has pretty high CO2 emissions. The others don't seem to be drastic in their emissions.

3. This is a plot of CO2 emissions per capita in the U.S, China, India, Canada, and Mexico over time. 
```{r, echo=FALSE}
new <- filter(gapminder, name == 'China'| name == 'India' | name == 'U.S' | name == 'Mexico' | name == 'Canada' | name == 'United States of America')

ggplot(new) +
  geom_point(mapping = aes(x=time, y=co2_PC, color = name))
```
The U.S has the highest CO2 emissions per capita out of all of the countries provided, while China is next and Mexico, India, and Canada are a lot lower on the graph.

4. This table shows the average CO2 emissions per capita across the continents.
```{r, echo=FALSE}
newg <- gapminder %>%
  select(region, co2_PC)
new2 <- na.omit(newg)
new2%>%
  group_by(region)%>%
  summarise(n = mean(co2_PC))
```


5. 
```{r, echo=FALSE}
plot <- gapminder%>%
  group_by(region, time)%>%
  filter(!is.na(co2_PC),
         time %in% c(1960, 2016),
         !is.na(region)) %>%
  summarize(avg_co2 = mean(co2_PC, na.re = TRUE))

ggplot(plot, aes(region, avg_co2, col = factor(time), fill = factor(time), na.re=TRUE)) +
  geom_bar(stat = "identity", position = "dodge", na.re=TRUE)

```

6. This table shows the three largest and three smallest CO2 emitters for each continent.
```{r, echo=FALSE}
newg<- gapminder%>%
 filter(time == 2010)%>%
  select(region, co2, name)
new2 <- na.omit(newg)
n <- new2%>%
  group_by(region)%>%
  arrange(co2)

top3<- n%>%
slice_tail(n=3)%>%
  arrange(desc(co2))

small3<-n%>%
  slice_head(n=3)
top3
small3
```
--->4. GDP per Capita

1. GDP per capita versus life expectancy in 1960
```{r, echo=FALSE}
gdp1960 <- gapminder%>%
  filter(time == 1960)

ggplot(data = gdp1960) +
  geom_point(mapping = aes(x=lifeExpectancy, y = GDP_PC, color = region, size = GDP_PC)) 
```
  
2. GDP per capita versus life expectancy in 2019   
```{r, echo=FALSE}
gdp1960 <- gapminder%>%
  filter(time == 2019)

ggplot(data = gdp1960) +
  geom_point(mapping = aes(x=lifeExpectancy, y = GDP_PC, color = region, size = GDP_PC)) 
```

3. The plots overall show the same pattern of Western countries having a higher life expectancy but the change between the years shows that every country is still increasing its life expectancy.

4. Yes the results fit.
```{r, echo=FALSE}
plot <- gapminder%>%
  group_by(region, time)%>%
  filter(!is.na(lifeExpectancy),
         time %in% c(1960, 2019),
         !is.na(region)) %>%
  summarize(avg_le = mean(lifeExpectancy, na.re = TRUE))
plot
```

5. 
```{r, echo=FALSE}
legrowth<- gapminder%>%
  select(region, time, lifeExpectancy)%>%
  group_by(region)%>%
  mutate(lediff = lifeExpectancy-lag(lifeExpectancy))%>%
  filter(time %in% c(1960, 2019))
print(legrowth)
```
  
6. 
```{r, echo=TRUE}
plot <- gapminder%>%
  group_by(GDP_PC, time)%>%
  filter(!is.na(GDP_PC),
         time %in% c(1960, 2019)) %>%
  summarize(avg_GDP = mean(GDP_PC, na.re = TRUE))

ggplot(plot, aes( time, GDP_PC, col = factor(time), fill = factor(time), na.re=TRUE)) +
  geom_bar(stat = "identity", position = "dodge", na.re=TRUE)

```

7.
```{r, echo=FALSE}
data <- gapminder %>%
  filter(time == 1960)

rank <- arrange(data, desc(lifeExpectancy))
us<- which(rank$name == "United States of America") +1
```
The United States is ranked `r us` in life expectancy in 1960.

```{r, echo=FALSE}
data <- gapminder %>%
  filter(time == 2019)

rank <- arrange(data, desc(lifeExpectancy))
us<- which(rank$name == "United States of America") +1
```
The United States is ranked `r us` in life expectancy in 2019.

8.
```{r, echo=FALSE}
data <- gapminder %>%
  filter(time == 1960)
num <- n_distinct(data$name)
rank <- arrange(data, desc(lifeExpectancy))

us<- which(rank$name == "United States of America")
us_rank <- us/num
```

The United States is ranked `r us_rank` relative to the number of countries with life expectancy data in 1960.

```{r, echo=FALSE}
data <- gapminder %>%
  filter(time == 2019)
num <- n_distinct(data$name)
rank <- arrange(data, desc(lifeExpectancy))

us<- which(rank$name == "United States of America")
us_rank <- us/num
```

The United States is ranked `r us_rank` relative to the number of countries with life expectancy data in 2019.
